msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-08 paper.tex

Recombination is implemented in \msprime\ using Hudson's algorithm, which
works backwards in time tracking the
effects of common ancestor and recombination
events~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
Common ancestor events combine the ancestral material of two lineages, and may
result in coalescences in the marginal trees. Recombination events
split the ancestral material for some lineage, creating two independent
lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
this process is much more efficient to simulate than either
methods based on the Ancestral Recombination Graph
formulation~\citep{griffiths1991two,griffiths1997ancestral}
or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
us to simulate recombination hotspots like \texttt{msHOT}~\citep{hellenthal2007mshot}
and empirical recombination maps similar to other
simulators~\citep[e.g.][]{shlyakhter2014cosi2}.

@@

The SMC~\citep{mcvean2005approximating} and
SMC'~\citep{marjoram2006fast} models are available
in \msprime\ 1.0. However, they are currently implemented using a
naive rejection sampling approach, and are somewhat slower
to simulate than the exact coalescent with recombination. These
models are therefore only appropriate for studying the SMC approximations
themselves; an important avenue for future work is to implement
efficient SMC-like models, perhaps based on the approaches used
in \texttt{cosi2}~\citep{shlyakhter2014cosi2}.

@@

We have also implemented gene
conversion following the model defined by~\cite{wiuf2000coalescent}
and used in \ms.

Gene conversion is particularly useful to model homologous recombination in
bacterial evolution. We compare the performance of \msprime\ with gene
conversion to \ms\ and the whole bacterial genome simulator
SimBac~\citep{brown2016simbac}. Further features are planned to model bacterial
evolution. In particular, we aim to include bacterial gene transfer and utilize
the succinct tree sequence structure to represent the resulting ancestral gene
transfer graph~\citep{baumdicker2014AGTG}.

@@

\section*{Results and Discussion}
\subsection*{Performance}
We compare \msprime\ with some existing simulators that have a subset of its
features.

\subsection*{Examples}
We show a few examples of running \msprime.

@@

>>>>

Recombination is implemented in \msprime\ using Hudson's algorithm, which
works backwards in time tracking the
effects of common ancestor and recombination
events~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
Common ancestor events combine the ancestral material of two lineages, and may
result in coalescences in the marginal trees. Recombination events
split the ancestral material for some lineage, creating two independent
lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
this process is much more efficient to simulate than either
methods based on the Ancestral Recombination Graph
formulation~\citep{griffiths1991two,griffiths1997ancestral}
or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
us to simulate recombination hotspots like \texttt{msHOT}~\citep{hellenthal2007mshot}
and empirical recombination maps similar to other
simulators~\citep[e.g.][]{shlyakhter2014cosi2}.
The implementation of recombination in \msprime\ is extensively validated
against analytical results~\citep{hudson1983properties,kaplan1985use}
and results from \ms.

@@

The SMC~\citep{mcvean2005approximating} and
SMC'~\citep{marjoram2006fast} models are available
in \msprime\ 1.0. However, they are currently implemented using a
naive rejection sampling approach, and are somewhat slower
to simulate than the exact coalescent with recombination. These
models are therefore only appropriate for studying the SMC approximations
themselves; an important avenue for future work is to implement
efficient SMC-like models, perhaps based on the approaches used
in \texttt{cosi2}~\citep{shlyakhter2014cosi2}, to facilitate
simulation of organisms such as \textit{Drosophila melagaster}.

@@

Gene conversion is \jkcomment{A couple of sentences describing gene
conversion, and what the model is. What other simulators simulate
gc? }
We have implemented this model of gene
conversion in \msprime\ 1.0, and validated the output against
\ms\ and analytical results~\citep{wiuf2000coalescent}.

\begin{figure}
\caption{\label{fig-gc-perf}Comparison of the running time
for E-coli simulations using \msprime\ and SimBac for varying
sample sizes. The parameters used are [] based on [references].}
\end{figure}

Gene conversion is particularly useful to model homologous recombination in
bacterial evolution.
We compare the performance of \msprime\ with gene conversion to
SimBac~\citep{brown2016simbac}, a simulator developed specifically to
simulate bacterial evolution. Although \msprime\ is missing some features
relevant to this case, it is far more efficient at simulating the basic
processes. An important avenue for future work
will be to add further features to model bacterial
evolution. In particular, we aim to include bacterial gene transfer and utilize
the succinct tree sequence structure to represent the resulting ancestral gene
transfer graph~\citep{baumdicker2014AGTG}.

@@

\section*{Results and Discussion}

@@

A significant advantage to simulating the coalescent with recombination
backwards in time using Hudson's algorithm rather than
left-to-right
approaches~\citep{wiuf1999recombination,chen2009fast,staab2014scrm}
is the ease with which additional processes can be integrated.
We have seen that processes such as gene conversion, complex
demography, instantaneous bottlenecks, multiple merger caolescents
and selective sweeps have been integrated seamlessly into
the simulations. Incorporating any of these extensions into
either the exact or approximate left-to-right processes would require
complex derivations. Because we are going backwards
in time, processes can be modelled as competing exponentials
using Gillespie's algorithm \jkcomment{I think?},
and therefore in principle we can
add as many extra processes as we like.
\jkcomment{so, backwards in time
sims in msprime is extensible for the future.}
